{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2022-11-15 Docs/manuscript.Rnw
cross ref ms and supp, model outputs
update manuscript
started bibtex

% required 
\usepackage{Sweave}
\usepackage{graphicx}
\usepackage{lipsum}                     % Dummytext % https://tex.stackexchange.com/questions/9796/how-to-add-todo-notes
\usepackage{xargs}                      % Use more than one optional parameter in a new commands
\usepackage[pdftex,dvipsnames]{xcolor}  % Coloured text etc.

@@

% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{November 2022}
\fancyhead[RO]{Manuscript}
% more optionals! %
\usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has

%%% end preambling. %%%

@@

\section{Introduction}
\begin{itemize}
\item Define spring event, fall event
% \item Background \unsure{What else?}
\item Research question (links to results)
\end{itemize}

@@

Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe (Fig. \ref{figure:springfall_latmat}). \unsure{Should I add climate overlap plots here too?}

@@

\bibliography{..//..//refs/ranges}
\section{Figures}

@@

>>>>

% required 
\usepackage{natbib}
\bibliographystyle{plainnat} %{unsrtnat}
\usepackage{xr-hyper}
\usepackage{hyperref}
\externaldocument[supplement-]{supplement}
\usepackage{booktabs,siunitx}
\usepackage{Sweave}
\usepackage{graphicx}
\usepackage{lipsum}                     % Dummytext % https://tex.stackexchange.com/questions/9796/how-to-add-todo-notes
\usepackage{xargs}                      % Use more than one optional parameter in a new commands
\usepackage[pdftex,dvipsnames]{xcolor}  % Coloured text etc.

@@

% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{November 2022}
\fancyhead[RO]{Manuscript}
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

@@

\section{Introduction}
\begin{itemize}
\item Define spring event, fall event
\item Background 
\item Research question (links to results)
\end{itemize}

\todo[inline]{Good background information.}
\citep{AitkenBemmels16}\\
"Geographic variation in trees has been investigated since the mid-18th century. Similar patterns of clinal variation have been observed along latitudinal and elevational gradients in common garden experiments for many temperate and boreal species."\\
"Scientists have studied relationships between tree populations and environmental characteristics of their provenances in common garden experiments for over 250 years."\\
\\
\citep{Alberto13}\\
"Evolutionary responses are required for tree populations to be able to track climate change. Results of 250 years of common garden experiments show that most forest trees have evolved local adaptation, as evidenced by the adaptive differentiation of populations in quantitative traits, reflecting environmental conditions of population origins."\\
"Thus, responding to climate change will likely require that the quantitative traits of populations again match their environments. We examine what kind of information is needed for evaluating the potential to respond, and what information is already available. We review the genetic models related to selection responses, and what is known currently about the genetic basis of the traits. We address special problems to be found at the range margins, and highlight the need for more modeling to understand specific issues at southern and northern margins. We need new common garden experiments for less known species. "\\
"Modeling work on the potential of populations and species to respond genetically to recent climate change is advancing (see Hoffmann \& Sgrò, 2011; Franks \& Hoffmann, 2012; Shaw \& Etterson, 2012 for recent reviews). "\\

(see e.g., Valladares et al., 2007; Caffarra et al., 2011; Hänninen \& Tanino, 2011)
"The plastic response of different traits (e.g., phenology in trees) to variation in climate is, however, often much more complex than in heuristic models of adaptation."\\

\citep{Loarie09}\\ 
"It is also now understood that the rate of adaptation required by climate change varies among geographic regions." \improvement{need to read again}\\

\citep{Nicotra10}\\ 
"The immediate responses via phenotypic plasticity have also been considered in the context of climate change." \improvement{need to read again}\\

\citep{Menzel99} \textbf{Growing season extended in Europe.}\improvement{good Europe background paper}\\
\citep{Menzel06} \textbf{European phenological response to climate change matches the warming pattern.}\improvement{another good Europe background paper}\\
\citep{Parmesan06}\\
"Trees exhibit a high degree of phenotypic plasticity with respect to climatic variation. Phenological shifts of bud flush in response to recent increases in temperatures have been widely documented ." \improvement{can cite more papers on this matter}

\citep{Gill15} \textbf{Changes in autumn senescence in northern hemisphere deciduous trees: a meta-analysis of autumn phenology studies.} \improvement{GREAT paper for us to base our writing on: similar meta-analysis but focuses on fall events}\\
"the effects of climate change on the phenology of vegetation have received increased attention over the past several decades as any factor that alters the timing of early growing season leaf-out and senescence has the potential to affect a variety of ecosystem properties. For example, the timing of leaf-out and senescence of deciduous plants has been shown to affect plant competition (Fridley, 2012), plant growth (Myneni et al., 1997) and ecosystem carbon uptake (Barichivich et al., 2012).\\

"While leaf-out has been shown to advance over the past century as a result of increasing air temperatures (Linderholm, 2006; Polgar and Primack, 2011), the relationship between temperature and leaf senescence remains less well understood (García-Plazaola et al., 2003; Richardson et al., 2013)."

"Thus, an extension of the growing season can contribute to reduced atmospheric CO2 concentrations due to enhanced carbon sequestration in terrestrial plants (Penuelas et al., 2009; Richardson et al., 2013). However, the increase in carbon uptake may be partially offset by increased rates of ecosystem respiration (Piao et al., 2008)"

"The lack of consistent relationships between air temperature and timing of leaf senescence suggests that autumn senescence may be influenced by a variety of factors that obscure its relationship with temperature."

\todo[inline]{Cite some papers about the documented trends}

tbc...

\todo[inline]{Alina says: I found AitkenBemmels2016's research question to be wonderfully phrased and think we can do something similar}
\citep{AitkenBemmels16}\\
"To what extent are patterns of local adaptation along climatic gradients similar among species within a given geographic region? If clines in phenotypic traits are similar, can average patterns of sympatric variation guide AGF or identify critical climatic variables as a first approximation for untested species? We reanalysed data from the literature on provenance trials in temperate tree species from western North America to address these questions. This is an ideal region of focus because of the many provenance trials conducted on native tree species that are important for forestry, and because its high topographic complexity means that turnover in climate occurs rapidly and heterogeneously over short geographic spaces. Clines along climate gradients are thus espcially likely to reflect local adaptation to climate, rather than non-adaptive phenomena resulting from population demographic history."

@@

Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe (Fig. \ref{figure:springfall_latmat} \& Supplement Table.\ref{supplement-table:model_sf_lat}). \unsure{Should I add climate overlap plots here too?}

@@

\citet{Keir11}

\section{Figures}

\bibliography{bibliography_local_adaptation.bib}

@@

% helping with abstract language
% Alberto16
% "On the basis of the patterns of quantitative variation for 19 adaptation-related traits studied in 59 tree species (mostly temperate and boreal species from the Northern hemisphere), we found that genetic differentiation between populations and clinal variation along environmental gradients were very common (respectively, 90% and 78% of cases)."

### 2022-11-15 Docs/manuscript.tex
cross ref ms and supp, model outputs
update manuscript
started bibtex

\section{Introduction}
\begin{itemize}
\item Define spring event, fall event
\item Background \unsure{What else?}
\item Research question (links to results)
\end{itemize}

>>>>

\section{Introduction}
\begin{itemize}
\item Define spring event, fall event
\item Background 
\item Research question (links to results)
\end{itemize}

\todo[inline]{Good background information.}
\citep{AitkenBemmels16}\\
"Geographic variation in trees has been investigated since the mid-18th century. Similar patterns of clinal variation have been observed along latitudinal and elevational gradients in common garden experiments for many temperate and boreal species."\\
"Scientists have studied relationships between tree populations and environmental characteristics of their provenances in common garden experiments for over 250 years."\\
\\
\citep{Alberto13}\\
"Evolutionary responses are required for tree populations to be able to track climate change. Results of 250 years of common garden experiments show that most forest trees have evolved local adaptation, as evidenced by the adaptive differentiation of populations in quantitative traits, reflecting environmental conditions of population origins."\\
"Thus, responding to climate change will likely require that the quantitative traits of populations again match their environments. We examine what kind of information is needed for evaluating the potential to respond, and what information is already available. We review the genetic models related to selection responses, and what is known currently about the genetic basis of the traits. We address special problems to be found at the range margins, and highlight the need for more modeling to understand specific issues at southern and northern margins. We need new common garden experiments for less known species. "\\
"Modeling work on the potential of populations and species to respond genetically to recent climate change is advancing (see Hoffmann \& Sgrò, 2011; Franks \& Hoffmann, 2012; Shaw \& Etterson, 2012 for recent reviews). "\\

(see e.g., Valladares et al., 2007; Caffarra et al., 2011; Hänninen \& Tanino, 2011)
"The plastic response of different traits (e.g., phenology in trees) to variation in climate is, however, often much more complex than in heuristic models of adaptation."\\

\citep{Loarie09}\\ 
"It is also now understood that the rate of adaptation required by climate change varies among geographic regions." \improvement{need to read again}\\

\citep{Nicotra10}\\ 
"The immediate responses via phenotypic plasticity have also been considered in the context of climate change." \improvement{need to read again}\\

\citep{Menzel99} \textbf{Growing season extended in Europe.}\improvement{good Europe background paper}\\
\citep{Menzel06} \textbf{European phenological response to climate change matches the warming pattern.}\improvement{another good Europe background paper}\\
\citep{Parmesan06}\\
"Trees exhibit a high degree of phenotypic plasticity with respect to climatic variation. Phenological shifts of bud flush in response to recent increases in temperatures have been widely documented ." \improvement{can cite more papers on this matter}

\citep{Gill15} \textbf{Changes in autumn senescence in northern hemisphere deciduous trees: a meta-analysis of autumn phenology studies.} \improvement{GREAT paper for us to base our writing on: similar meta-analysis but focuses on fall events}\\
"the effects of climate change on the phenology of vegetation have received increased attention over the past several decades as any factor that alters the timing of early growing season leaf-out and senescence has the potential to affect a variety of ecosystem properties. For example, the timing of leaf-out and senescence of deciduous plants has been shown to affect plant competition (Fridley, 2012), plant growth (Myneni et al., 1997) and ecosystem carbon uptake (Barichivich et al., 2012).\\

"While leaf-out has been shown to advance over the past century as a result of increasing air temperatures (Linderholm, 2006; Polgar and Primack, 2011), the relationship between temperature and leaf senescence remains less well understood (García-Plazaola et al., 2003; Richardson et al., 2013)."

"Thus, an extension of the growing season can contribute to reduced atmospheric CO2 concentrations due to enhanced carbon sequestration in terrestrial plants (Penuelas et al., 2009; Richardson et al., 2013). However, the increase in carbon uptake may be partially offset by increased rates of ecosystem respiration (Piao et al., 2008)"

"The lack of consistent relationships between air temperature and timing of leaf senescence suggests that autumn senescence may be influenced by a variety of factors that obscure its relationship with temperature."

\todo[inline]{Cite some papers about the documented trends}

tbc...

\todo[inline]{Alina says: I found AitkenBemmels2016's research question to be wonderfully phrased and think we can do something similar}
\citep{AitkenBemmels16}\\
"To what extent are patterns of local adaptation along climatic gradients similar among species within a given geographic region? If clines in phenotypic traits are similar, can average patterns of sympatric variation guide AGF or identify critical climatic variables as a first approximation for untested species? We reanalysed data from the literature on provenance trials in temperate tree species from western North America to address these questions. This is an ideal region of focus because of the many provenance trials conducted on native tree species that are important for forestry, and because its high topographic complexity means that turnover in climate occurs rapidly and heterogeneously over short geographic spaces. Clines along climate gradients are thus espcially likely to reflect local adaptation to climate, rather than non-adaptive phenomena resulting from population demographic history."
